\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-03-12 papers/methods/paper.tex
Anais feedback 2

We test four models: Flat $\Lambda$CDM, Flat $w$CDM, $\Lambda$CDM and Flat $w$CDM with a prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, with the latter included to allow sensitive tests on bias for $w$. To achieve statistical precision, we fit 100 realisations of supernovae datasets. Cosmological parameters are recovered without significant bias. Combined posterior surfaces of all 100 realisations fits for $\Lambda$CDM are shown in Figure \ref{fig:simple_ol} and fits for Flat $w$CDM are shown in Figure \ref{fig:simple_w}. By utilising the Stan framework and several efficient parametrisations (discussed further in Appendix \ref{app:optimisations}), fits to these simulations of 2000 supernovae take only on order of a single CPU-hour to run.

@@

Early analyses often treated intrinsic dispersion simply as scatter in the underlying absolute magnitude of the underlying population, but recent analyses require more a more sophisticated approach. In our development of this model and tests of intrinsic dispersion, we analyse the effects of two different scatter models. The first model is the \citet[][hereafter denoted {\gten}]{Guy2010} scatter model, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second model, denoted the {\celeven} model, is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation.

Simulations (using the SNANA package) follow the observational schedule and observing conditions for the DES and LowZ surveys. In addition to the improvements in the scatter models over the simple data, we also include peculiar velocities for the LowZ sample, and now also include our full treatment of systematics. Our simulated populations are sourced from \citet[][hereafter {\sk}]{Scolnic2016}. The selection effects observed by comparing the generated supernovae to those that pass our simulated cuts and were successfully `observed' are shown in Figure~\ref{fig:sf_snana}, and it is from this simulation that our analytic determination of the selection functions for the LowZ and DES survey are based. We note that there was no significant difference in the selection effect of Malmquist bias between using the {\gten} or {\celeven} scatter model compared to the fit uncertainty.

@@

Each realisation of cosmology simulated contains 137 LowZ supernovae, and 204 DES-like supernovae, such that the uncertainties found when combining chains is representative of the uncertainty in the final DES spectroscopic analysis. As our primary focus is upon Dark Energy, we now focus specifically on the Flat $w$CDM model with matter prior.

@@

Cosmological bias is detected with $\approx 2.7\sigma$ significance for the {\celeven} model and $\approx 1.5\sigma$ significance for the {\gten} model when fitting both systematics and statistics. For the statistics only fits, the respective bias detection levels are $0.9\sigma$ and $5.5\sigma$. We investigate the source of the {\celeven} bias and find its source to be bias in the observed summary statistics, in addition to incorrect reported uncertainty on those summary statistics. By replacing fully simulated observed $m_B$, $x_1$ and $c$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty is incorrect, non-Gaussian and biased. Our current model and techniques can quantify the effect of different scatter models, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model.

@@

In this paper we have outlined the creation of a hierarchical Bayesian model for supernova cosmology. The model takes into account selection effects and their uncertainty, fits underlying populations and standardisation parameters, incorporates unexplained dispersion from intrinsic scatter colour smearing and incorporates uncertainty from peculiar velocity uncertainty, calibration uncertainty, dust uncertainty and SALT2 model uncertainty. Furthermore, these uncertainties, being explicitly parametrised in our model, are captured will covariance intact, an improvement on many previous methods. The model has been optimised to allow for hundreds of supernovae to be modelled fully with latent parameters and run in under an hour of CPU time and scales linearly with the number of supernovae, as opposed to polynomial complexity of matrix inversion.

>>>>

We test four models: Flat $\Lambda$CDM, Flat $w$CDM, $\Lambda$CDM, and Flat $w$CDM with a prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, with the latter included to allow sensitive tests on bias for $w$. To achieve statistical precision, we fit 100 realisations of supernovae datasets. Cosmological parameters are recovered without significant bias. Combined posterior surfaces of all 100 realisations fits for $\Lambda$CDM are shown in Figure \ref{fig:simple_ol} and fits for Flat $w$CDM are shown in Figure \ref{fig:simple_w}. By utilising the Stan framework and several efficient parametrisations (discussed further in Appendix \ref{app:optimisations}), fits to these simulations of 2000 supernovae take only on order of a single CPU-hour to run.

@@

Early analyses often treated intrinsic dispersion simply as scatter in the underlying absolute magnitude of the underlying population, but recent analyses require a more sophisticated approach. In our development of this model and tests of intrinsic dispersion, we analyse the effects of two different scatter models. The first is the \citet[][hereafter denoted {\gten}]{Guy2010} scatter model, which models intrinsic scatter with a 70\% contribution from coherent variation and 30\% from chromatic variation. The second, denoted the {\celeven} model, is sourced from \citet{Chotard2011} and has variation with 25\% contribution from coherent scatter and 75\% from chromatic variation.

Simulations (using the SNANA package) follow the observational schedule and observing conditions for the DES and LowZ surveys. In addition to the improvements in the scatter models over the simple data, we also include peculiar velocities for the LowZ sample, and our full treatment of systematics. Our simulated populations are sourced from \citet[][hereafter {\sk}]{Scolnic2016}. The selection effects were quantified by comparing all the generated supernovae to those that pass our simulated cuts, as shown in Figure~\ref{fig:sf_snana}. It is from this simulation that our analytic determination of the selection functions for the LowZ and DES survey are based. We note that there was no significant difference in the selection effect of Malmquist bias between using the {\gten} or {\celeven} scatter model compared to the fit uncertainty.

@@

Each realisation of simulated cosmology contains 137 LowZ supernovae, and 204 DES-like supernovae, such that the uncertainties found when combining chains is representative of the uncertainty in the final DES spectroscopic analysis. As our primary focus is Dark Energy, we now focus specifically on the Flat $w$CDM model with matter prior.

@@

Cosmological bias is detected with $\approx 2.7\sigma$ significance for the {\celeven} model and $\approx 1.5\sigma$ significance for the {\gten} model when fitting both systematics and statistics. For the statistics only fits, the respective bias detection levels are $0.9\sigma$ and $5.5\sigma$. We investigate the {\celeven} bias and find its source to be a bias in the observed summary statistics, in addition to incorrect reported uncertainty on the summary statistics. This is confirmed by replacing the fully simulated observed $m_B$, $x_1$ and $c$ with random numbers drawn from a true Gaussian centered on the simulated SALT2 $m_B$, $x_1$ and $c$ values with covariance as reported by initial light curve fits. With this test, both the {\gten} and {\celeven} fits recover $w=-1.00$ exactly. From this, the main challenge of improving our methodology is to handle the fact that observational uncertainty is incorrect, non-Gaussian and biased. Our current model and techniques can quantify the effect of different scatter models, but being unable to constrain the `correct' (simulated) scatter model in our model fit means we cannot fully correct for the bias introduced an unknown scatter model.

@@

In this paper we have outlined the creation of a hierarchical Bayesian model for supernova cosmology. The model takes into account selection effects and their uncertainty, fits underlying populations and standardisation parameters, incorporates unexplained dispersion from intrinsic scatter colour smearing and incorporates uncertainty from peculiar velocities, survey calibration, HST calibration, dust, a potential global redshift offset, and SALT2 model uncertainty. Furthermore, our uncertainties in standardisation, population, mass-step and more, being explicitly parametrised in our model, are captured will covariance intact, an improvement on many previous methods. The model has been optimised to allow for hundreds of supernovae to be modelled fully with latent parameters and run in under an hour of CPU time and scales linearly with the number of supernovae, as opposed to polynomial complexity of matrix inversion of other methods.
